\sys: Property-Guided Program Regeneration\\{\vspace{1em}\large Technical University of Denmark, MSc Thesis {\\\vspace{-1em}} Advisors: Christian Gram Kahlauge and Nikos Vasilakis

### 2025-07-18 docs/paper/draft.tex
Tweaks
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

\begin{abstract}
  Software evolution tasks---like security hardening, porting critical components to safer 
  languages, or refactoring opaque legacy code---are brittle and time-consuming.
  Large language models (LLMs) have shown promise in automating such tasks,
  but their outputs are shaped by surface cues, lack
  semantic guarantees, and their usage often boil down to brittle prompting strategies.
This work presents \sys, a system for controllable program regeneration that
  combines declarative property constraints with LLM-guided synthesis. Unlike
  prior approaches, \sys treats the input program as an untrusted artifact and
  explicitly projects it into a minimal set of verifiable properties, such as
  input/output behavior, execution traces, and API signatures. These properties
  serve as constraints during synthesis and are checked post hoc to ensure
  correctness.
\sys is evaluated on a diverse set of program transformation tasks
including security hardening, language translation, idiomatic rewriting, de-obfuscation,
and modular refactoring showing strong results, even on tasks where state-of-the-art LLMs fail.
\end{abstract}
\maketitle

@@

I present \sys, a system for \emph{property-guided program regeneration}.
\sys allows users
to express transformation goals declaratively, as logical constraints over
program properties such as input/output behavior, side-effect freedom, target
language, trace equivalence, or modular structure.
As shown in \cref{fig:overview}, given a source program and
a query, \sys extracts a minimal set of relevant properties, guided by a
logic engine and a domain-specific knowledge base, and synthesizes a new
program that satisfies the desired constraints.

@@

\heading{Answer Set Programming}
Answer Set Programming (ASP) \cite{Gelfond_2000, Eiter_2009} is a problem-solving paradigm with roots in logic programming and non-monotonic reasoning.
The programming 
model of ASP is one where the programmer models
the problem domain, with the solution being handled by a solver program.
Programming using this paradigm is done in a family of languages sometimes called \textit{AnsProlog} \cite{Gelfond_2002}.
This work uses the input language of \textit{Clingo} \cite{DBLP:journals/corr/GebserKKS14}, which 
is a high-performance integrated solver with a large collection of libraries and bindings helpful in integrating it with external tools.

@@

To validate that a regenerated program satisfies the intended goals, \sys includes an explicit verification phase that checks whether the output adheres to the enforced properties in the synthesis plan. 
Verification accepts only programs that meet the plan's constraints and provides feedback for refining or retrying those that do not.

@@

Examples include:
\begin{itemize}
  \item For $\texttt{no\_network}(f)$, the checker confirms the absence of known network APIs or imports.
  \item For $\texttt{signature}(f)$, the checker matches the regenerated function's name, parameters, and arity against the extracted signature.
  \item For $\texttt{pure}(f)$, the checker examines control flow and side-effecting operations to detect possible violations.
\end{itemize}

Checkers are modular and contributed by plugins. Their combined results determine whether a candidate is considered valid under the current plan.
For minimization and maximization goals, the verification phase uses the less-than or greater-than operators as are defined internally by the plugin verifying the given property.
Verification in this case checks if the property's value has decreased or increased compared 
to the previous regeneration attempt.
If the value remains unchanged for two consecutive attempts, its considered satisfied.

@@

The \sys evaluation is along three axes:

\begin{itemize}
  \item[\textbf{Q1}] \textbf{Correctness and Regeneration Quality}: Does \sys preserve program behavior and produce high-quality programs?
  \item[\textbf{Q2}] \textbf{LLM Baseline}: How does \sys compare to naively prompting a large language model?
  \item[\textbf{Q3}] \textbf{Planning Efficiency \& Precision}: Does minimizing extracted properties help regeneration succeed faster and more reliably?
\end{itemize}

\section{Benchmarks \& Methodology}

The \sys evaluation comprises of six benchmark suites, summarized in \cref{tab:benchmarks}.
Each task consists of an input program, a \sys query expressing desired
properties, and a test suite or functional oracle.
Each benchmark is executed three times. A regeneration is considered overall successful 
if it is correct, satisfies the query, and passes all tests all three times.
The Rosetta Code~\cite{rosettacode} benchmarks include four programming tasks written in four different
languages each (Python, JavaScript, C, and Haskell). 
These include classic programs that do not require any language-specific features to implement.
These are: the classic \emph{Hello, World!} program as found in most programming tutorials,
a program that applies the ROT-13 cipher to a string,
a program that computes the median of a list of numbers,
and a program that reads a file and prints its contents to the console line-by-line.
The npm benchmark includes ten popular (100.000+ weekly downloads) npm packages that implement utility
functionality such as string manipulation, data structure transformations, and math.
Similarly, the PyPi utilities benchmark includes three functions sourced from PyPi that implement similar functionality.
The SSCA dataset includes three sneaky supply-chain attacks that obfuscate malicious behavior in otherwise innocuous-looking code.
The attack activates under specific environment conditions, such as the presence of a specific environment variable or specific inputs
getting passed to the program. The attacks involve private key stealing~\cite{ohm2020backstabber}, Bitcoin private key theft~\cite{ev:eurosec:2022}, and an SSH backdoor installation~\cite{copeland2019frightening}
and occur while the program still performs its advertised functionality.
The IOCCC benchmark includes two programs from the International Obfuscated C Code Contest (IOCCC)~\cite{ioccc}.
These programs are intentionally obfuscated, using advanced techniques such as bit manipulation, undefined behavior,
and exploitation of the C preprocessor and the lax parsing rules of the C language.
These programs perform mathematical computations like Mersenne primality testing and adding two numbers.
Finally, the microbenchmarks include three programs that implement the regeneration tasks discussed in \cref{sec:example}.

@@

\sys successfully regenerates all programs in the Rosetta Code benchmark suite,
having verified manually that the output is functionally equivalent
to the original, using both the original and translated reference
implementations as guides.
\sys regenerates correctly all npm and PyPi utilities, it does not introduce and side-effectful or malicious behavior.
The synthesized programs reveal interesting insights and patterns.
\sys-generated code often lacks the optimizations commonly
implemented by library developers, such as early exits or caching.
For instance, the \texttt{left-pad} library, pre-computes padding for strings below a certain length
and returns it directly.
In contrast, \sys tends to produce straightforward
implementations without attempting similar optimizations.
\sys also omits domain-specific knowledge:
\sys-generated code frequently fails to account for all edge cases when they involve language-specific features (\eg \ttt{is-nan}).
% \sys has weak performance on libraries that involve complex regexes or perform non-trivial math.
For the SSCA benchmarks, \sys successfully regenerates all three malicious libraries, removing the malicious behavior while preserving the original functionality.
For the flatmap-stream attack, \sys uses domain-specific logic (encoded in its knowledge-base by the \ttt{js-stream} plugin) to generate a function that produces a node \ttt{Stream} object.
For the IOCCC benchmarks, \sys regenerates the \ttt{adder} program, which is a simple program that adds two numbers,
it also regenerates \ttt{mersenne} in regards that it correctly performs a naive Mersenne primality testing, but it does not follow the non-obvious output scheme of the original program, which returns a non-zero hexadecimal number when $2^p - 1$ ($p$ being the input number) is not prime.
For the microbenchmarks, \sys correctly regenerates the \ttt{flatmap-stream} program, producing a safe Haskell version that is idiomatic and preserves the original functionality.
However, this version does not match exactly the semantics of the original one, as it operates over infinitely nested lists, rather than only one level deep ones.
The \ttt{Q\_rsqrt} program includes the obvious, idiomatic computation of the inverse square root ($1/\sqrt{x}$) rather than the original bit-level manipulation.
Finally, \sys successfully decomposes the \ttt{rigidDB} program into two
functions, \ttt{setupDatabase} and \ttt{getAlbumsByArtist}, which is a
modularization that preserves the original functionality and produce equivalent I/O and SQL traces.

@@

\sys consistently outperforms a \gptmodel baseline,
overcoming the inherent weaknesses of a purely prompt-based approach.

\subsection{Q3: Planning Efficiency \& Precision}

To evaluate the impact of \sys's property planning, we compare it against a \gptmodel baseline that performs prompt-only rewriting without explicit access to guiding properties.

We consider two criteria for each regenerated program:

\begin{itemize}
  \item \textbf{Wanted Property Satisfaction:} Does the output program satisfy all positive properties declared in the query (\eg \texttt{pure}, \texttt{language(haskell)})?
  \item \textbf{Unwanted Property Avoidance:} Does the output avoid introducing disallowed behaviors (\eg side effects, global state, unsupported language features)?
\end{itemize}

\sys consistently satisfies more of the declared properties while avoiding
unintended behaviors.
Following is a qualitative analysis of the results.
In the \texttt{flatmap-stream} benchmark,
\gptmodel often inclduded malicious side effects when the code appeared innocuous,
whereas \sys generated pure rewrites by treating the source as
untrusted and grounding regeneration in I/O behavior alone.
Across all benchmarks, the planning step completes in under 50ms on average.
Similarly, in the \ttt{Q\_rsqrt} benchmark, \gptmodel often produces bit-level
manipulations that are not idiomatic in Haskell, while \sys generates a simple
and idiomatic Haskell function that computes the inverse square root.
This is guided by the fact the \sys extracts and provides to its backend model only the function signature 
and length of the program, which it minimizes, without including the original source code,
which as shown in the experimental results makes the model over-fit its regeneration to it.
\sys's I/O pair extraction also provides confidence that 
the regenerated program has the same client-visible behavior as the original one, while 
prompting a model like \gptmodel would always require manual inspection or the presence 
of developer tests to verify the correctness of the regeneration.

@@

\heading{Logic programming and constraint solving}
Logic programming has long served as a foundation for reasoning in programming
tools, offering a declarative model for expressing properties and constraints.
Answer Set Programming (ASP), in particular, provides a rich framework for
expressing defaults, exceptions, and optimization via stable model semantics
\cite{Gelfond_2000, Gelfond_2002, Eiter_2009}. 
ASP has been used for
declarative program analysis \cite{benton2007interactive}, automated planning
\cite{nguyen2020explainable, son2022answersetplanningsurvey}, and knowledge
representation in verification and synthesis pipelines.

@@

\section{Conclusion}
I introduced \sys, a system for property-guided program regeneration
that bridges the flexibility of large language models with the precision of
logic-driven planning.
By allowing developers to declaratively specify
transformation goals as logical constraints, \sys elevates LLM-assisted program
transformations to a controllable, interpretable, and iterative synthesis-verification
process.
Evaluation demonstrates that \sys produces high-quality regenerations
across diverse tasks, outperforming naive prompting baselines in correctness,
controllability, and resilience to adversarial or obfuscated inputs.

@@

\heading{Ethical and legal considerations}
While this work has applied \sys exclusively to open-source software, its
capabilities raise ethical and legal questions. Regenerating
proprietary software, de-obfuscating binaries, or reimplementing licensed
systems into equivalent ones with more permissive licensing 
could challenge existing norms and
regulations. 
It is important to consider
safeguards and responsible disclosure practices towards ethical program regeneration.

@@

\section{Evaluation Samples}
This section provides a selection of samples from the evaluation of \sys.
These instances provide interesting insights into the capabilities of \sys and its shortcomings.
The samples also provide concrete comparison between \sys and a naive LLM-based approach.

>>>>

\begin{abstract}
  Software evolution tasks---like security hardening, porting critical components to safer 
  languages, or refactoring opaque legacy code---are brittle and time-consuming.
  Large language models (LLMs) have shown promise in automating such tasks,
  but their outputs are shaped by surface cues, lack
  semantic guarantees, and their usage often boil down to fragile prompting strategies.
This work presents \sys, a system for controllable program regeneration that
  combines declarative property constraints with LLM-guided synthesis. Unlike
  prior approaches, \sys treats the input program as an untrusted artifact and
  explicitly projects it into a minimal set of verifiable properties, such as
  input/output behavior, execution traces, and API signatures. These properties
  serve as constraints during synthesis, are checked post hoc to ensure
  correctness, and can be mutated to allow for flexible program transformation through regeneration.
\sys is evaluated on a diverse set of program transformation tasks
including security hardening, language translation, idiomatic rewriting, de-obfuscation,
and modular refactoring showing strong results, even on tasks where state-of-the-art LLMs fail.
\end{abstract}
\maketitle

@@

This work presents \sys, a system for \emph{property-guided program regeneration}.
\sys allows users
to express transformation goals declaratively, as logical constraints over
program properties such as input/output behavior, side-effect freedom, target
language, trace equivalence, or modular structure.
As shown in \cref{fig:overview}, given a source program and
a query, \sys extracts a minimal set of relevant properties, guided by a
logic engine and a domain-specific knowledge base, and synthesizes a new
program that satisfies the desired constraints.

@@

\heading{Answer Set Programming}
ASP \cite{Gelfond_2000, Eiter_2009} is a problem-solving paradigm with roots in logic programming and non-monotonic reasoning.
The programming 
model of ASP is one where the programmer models
the problem domain, with the solution being handled by a solver program.
Programming using this paradigm is done in a family of languages sometimes called \textit{AnsProlog} \cite{Gelfond_2002}.
This work uses the input language of \textit{Clingo} \cite{DBLP:journals/corr/GebserKKS14}, which 
is a high-performance integrated solver with a large collection of libraries and bindings helpful in integrating it with external tools.

@@

\sys includes an explicit verification phase that checks whether the output adheres to the enforced properties in the synthesis plan. 
Verification accepts only programs that meet the plan's constraints and
provides feedback for refining or retrying those that do not.

@@

Examples include:
\begin{itemize}
  \item For $\texttt{no\_network}(f)$, the checker confirms the absence of known network APIs or imports.
  \item For $\texttt{signature}(f)$, the checker matches the regenerated function's name, parameters, and arity against the extracted signature.
\end{itemize}

Checkers are modular and contributed by plugins.
Their combined results determine whether a candidate is considered valid under the current plan.
For minimization and maximization goals, the verification phase uses the less-than or greater-than operators as are defined internally by the plugin verifying the given property.
Verification in this case checks if the property's value has decreased or increased compared 
to the previous regeneration attempt.
If the value remains unchanged for two consecutive attempts, its considered satisfied.
Its important that certain properties that are not easily verifiable, such as \ttt{pure}, 
do not correspond to a specific verifier, but rather influence what kind of propoerties can be extracted 
otherwise to make sure they are not violated during regeneration.
For this example, a system call trace of the program will not be provided to the LLM instance, 
as that could encode side-effects present in the original program.

@@

\sys's evaluation aims to answer two key questions:

\begin{itemize}
  \item[\textbf{Q1}] \textbf{Correctness and Regeneration Quality}: Does \sys preserve program behavior and produce high-quality programs?
  \item[\textbf{Q2}] \textbf{LLM Baseline}: How does \sys compare to naively prompting a large language model?
\end{itemize}

\subsection{Benchmarks \& Methodology}

The \sys evaluation comprises of six benchmark suites, summarized in \cref{tab:benchmarks}.
Each task consists of an input program, a \sys query expressing desired
properties, and a test suite or functional oracle.
Each benchmark is executed three times. A regeneration is considered overall successful 
if it is correct, satisfies the query, and passes all tests all three times.
The specific benchmark programs were selected based on a combination of their ubiquity in programming practise, their
popularity in programming culture, and their tractability given the implemented \sys plugins.
The Rosetta Code~\cite{rosettacode} benchmarks include four programming tasks written in four different
languages each (Python, JavaScript, C, and Haskell). 
These include classic programs that do not require any language-specific features to implement.
These are: the classic \emph{Hello, World!} program as found in most programming tutorials,
a program that applies the ROT-13 cipher to a string,
a program that computes the median of a list of numbers,
and a program that reads a file and prints its contents to the console line-by-line.
The npm benchmark includes ten popular (100.000+ weekly downloads) npm packages that implement utility
functionality such as string manipulation, data structure transformations, and math.
Similarly, the PyPi utilities benchmark includes three functions sourced from PyPi that implement similar functionality.
The SSCA dataset includes three sneaky supply-chain attacks that obfuscate malicious behavior in otherwise innocuous-looking code.
The attack activates under specific environment conditions, such as the presence of a specific environment variable or specific inputs
getting passed to the program. The attacks involve private key stealing~\cite{ohm2020backstabber}, Bitcoin private key theft~\cite{ev:eurosec:2022}, and an SSH backdoor installation~\cite{copeland2019frightening}
and occur while the program still performs its advertised functionality.
The IOCCC benchmark includes two programs from the International Obfuscated C Code Contest (IOCCC)~\cite{ioccc}.
These programs are intentionally obfuscated, using advanced techniques such as bit manipulation, undefined behavior,
and exploitation of the C preprocessor and the lax parsing rules of the C language.
These programs perform mathematical computations like Mersenne primality testing and adding two numbers.
Finally, the microbenchmarks include three programs that implement the regeneration tasks discussed in \cref{sec:example}.

@@

\sys successfully regenerates all programs in the Rosetta Code benchmark suite,
having verified manually that the output is functionally equivalent
to the original, using both the original and translated reference
implementations as guides.
\sys regenerates correctly all npm and PyPi utilities, it does not introduce and side-effectful or malicious behavior.
The synthesized programs reveal interesting insights and patterns.
\sys-generated code often lacks the optimizations commonly
implemented by library developers, such as early exits or caching.
For instance, the \texttt{left-pad} library, pre-computes padding for strings below a certain length
and returns it directly.
In contrast, \sys tends to produce straightforward
implementations without attempting similar optimizations.
\sys also omits domain-specific knowledge:
\sys-generated code frequently fails to account for all edge cases when they involve language-specific features (\eg \ttt{is-nan}).
% \sys has weak performance on libraries that involve complex regexes or perform non-trivial math.
For the SSCA benchmarks, \sys successfully regenerates all three malicious libraries, removing the malicious behavior while preserving the original functionality.
For the flatmap-stream attack, \sys uses domain-specific logic (encoded in its knowledge-base by the \ttt{js-stream} plugin) to generate a function that produces a node \ttt{Stream} object.
For the IOCCC benchmarks, \sys regenerates the \ttt{adder} program, which is a simple program that adds two numbers,
it also regenerates \ttt{mersenne} in regards that it correctly performs a naive Mersenne primality testing, but it does not follow the non-obvious output scheme of the original program, which returns a non-zero hexadecimal number when $2^p - 1$ ($p$ being the input number) is not prime.
For the microbenchmarks, \sys correctly regenerates the \ttt{flatmap-stream} program, producing a safe Haskell version that is idiomatic and preserves the original functionality.
However, this version does not match exactly the semantics of the original one, as it operates over infinitely nested lists, rather than only one level deep ones.
Thus, a $.5$ score in correctness is given.
The \ttt{Q\_rsqrt} program includes the obvious, idiomatic computation of the inverse square root ($1/\sqrt{x}$) rather than the original bit-level manipulation.
Finally, \sys successfully decomposes the \ttt{rigidDB} program into two
functions, \ttt{setupDatabase} and \ttt{getAlbumsByArtist}, which is a
modularization that preserves the original functionality and produce equivalent I/O and SQL traces.

@@

\sys consistently outperforms a \gptmodel baseline,
overcoming the inherent weaknesses of a purely prompt-based approach.
In-depth examples of interesting evaluation results are discussed in \cref{sec:eval-samples}.

@@

\heading{Logic programming and constraint solving}
Logic programming has long served as a foundation for reasoning in programming
tools, offering a declarative model for expressing properties and constraints.
ASP, in particular, provides a rich framework for
expressing defaults, exceptions, and optimization via stable model semantics
\cite{Gelfond_2000, Gelfond_2002, Eiter_2009}. 
ASP has been used for
declarative program analysis \cite{benton2007interactive}, automated planning
\cite{nguyen2020explainable, son2022answersetplanningsurvey}, and knowledge
representation in verification and synthesis pipelines.

@@

\section{Conclusion}
This work introduced \sys, a system for property-guided program regeneration
that bridges the flexibility of large language models with the precision of
logic-driven planning.
By allowing developers to declaratively specify
transformation goals as logical constraints, \sys elevates LLM-assisted program
transformations to a controllable, interpretable, and iterative synthesis-verification
process.
Evaluation demonstrates that \sys produces high-quality regenerations
across diverse tasks, outperforming naive prompting baselines in correctness,
controllability, and resilience to adversarial or obfuscated inputs.

@@

\heading{Ethical and legal considerations}
While this work has applied \sys exclusively to open-source software, its
capabilities raise ethical and legal questions.
Regenerating
proprietary software, de-obfuscating binaries, or reimplementing licensed
systems into equivalent ones with more permissive licensing 
could challenge existing norms and regulations. 
It is important to consider
safeguards and responsible disclosure practices towards ethical program regeneration.
Solutions to similar problems have been

@@

\section{Evaluation Samples}
\label{sec:eval-samples}
This section provides a selection of samples from the evaluation of \sys.
These instances provide interesting insights into the capabilities of \sys and its shortcomings.
The samples also provide concrete comparison between \sys and a naive LLM-based approach.
